``` markdown
---
title: "Credit Card Transaction Analysis"
author: "Hamza Shaikh"
date: "2025-5-31"
output: html_document
---
```

# Credit Card Transaction Analysis

## Load Libraries

We start by loading the necessary libraries for our analysis.

```{r load-libraries}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(forecast)
library(randomForest)
library(tseries)
library(reshape2)
library(dplyr)
```

## Load Dataset

Load the dataset into R.

```{r load-dataset}
df <- read.csv("A:/GithuB/Credit Card Spending Analysis/data/Credit card transactions.csv", stringsAsFactors = FALSE)
head(df)
```

## Data Cleaning

We handle missing values and parse timestamps.

```{r data-cleaning}
df$Date <- dmy(df$Date)
df <- na.omit(df)
```

## Exploratory Data Analysis (EDA)

Perform EDA to understand the data structure and summary.

```{r eda}
summary(df)
str(df)
```

## Add Date Features

Extract year, month, and weekday from the date.

```{r add-date-features}
df$Year <- year(df$Date)
df$Month <- month(df$Date)
df$Weekday <- weekdays(df$Date)
```

## Aggregations

Compute total spend by city.

```{r aggregations}
total_spend_city <- df %>% group_by(City) %>% summarise(TotalSpend = sum(Amount)) %>% arrange(desc(TotalSpend))
print(total_spend_city)
```

## Transaction Count by Card Type

Count transactions by card type.

```{r transaction-count-by-card-type}
card_counts <- df %>% group_by(Card.Type) %>% summarise(Count = n())
print(card_counts)
```

## Histogram of Amount

Plot the distribution of transaction amounts.

```{r histogram-of-amount}
ggplot(df, aes(x=Amount)) +
  geom_histogram(bins=50, fill='skyblue', color='black') +
  ggtitle("Transaction Amount Distribution")
```

## Time Series Aggregation

Aggregate data by month to analyze spending trends.

```{r time-series-aggregation}
monthly_spend <- df %>% group_by(YearMonth = floor_date(Date, "month")) %>%
  summarise(MonthlyAmount = sum(Amount))

print(monthly_spend)
```

## Plot Monthly Spend

Visualize the monthly spending trend.

```{r plot-monthly-spend}
ggplot(monthly_spend, aes(x=YearMonth, y=MonthlyAmount)) +
  geom_line(color='darkgreen') +
  ggtitle("Monthly Spending Trend")
```

## Anomaly Detection using Z-score

Detect anomalies in transaction amounts using Z-score.

```{r anomaly-detection}
df$z_score <- scale(df$Amount)
df$Anomaly <- abs(df$z_score) > 3
```

## Plot Anomalies

Visualize anomalies in transactions.

```{r plot-anomalies}
ggplot(df, aes(x=Date, y=Amount, color=Anomaly)) +
  geom_point(alpha=0.7) +
  ggtitle("Anomalies in Transactions")
```

## Time Series Forecasting

Use ARIMA for time series forecasting.

```{r time-series-forecasting}
ts_data <- ts(monthly_spend$MonthlyAmount, start = c(2014,1), frequency = 12)
model <- auto.arima(ts_data)
forecasted <- forecast(model, h=6)
```

## Plot Forecast

Visualize the forecasted monthly spend.

```{r plot-forecast}
autoplot(forecasted) + ggtitle("ARIMA Forecast for Monthly Spend")
```

\`\`\`
